---
title: "lockdown risk analysis"
author: "Tiffany Wan"
date: "2026-04-17"
output: html_document
---



## load library

```{r load library}
library(pheatmap)
library(tidyverse)
library(readxl)
library(data.table)
library(janitor)
library(viridis)
library(randomcoloR)
library(RColorBrewer)
library(ape)
library(ggtree)
library(compareDF)
library(data.table)
library(tidyverse)
library(tidytext)
library(randomcoloR)
library(countrycode)
library(phytools)
library(aplot)
library(ggnewscale)
library(pscl)
library(magrittr)
library(nnet)
library(ggupset)
#library(olsrr)
library(car)
library(hrbrthemes)
library(plotly)
library(table1)
library(tableone)
library(scatterpie)
library(aplot)
library(igraph)
library(tidygraph)
library(ggraph)
library(lmerTest)

```



# variant call results including gubbins and MLtree files
```{r, echo=FALSE,fig.show='hide'}
# read variant calling results
#131
Ecoli_131 <- read.FASTA("/Users/tifwan/Desktop/flux_mount_gl/Project_IMPALA/old_Sequence_data/output_files/2024_01_12_IMPALA_variant_calling/Ecoli/EcoliST131/2024_02_29_15_04_03_core_results/gubbins/2024_02_29_15_04_03_Ecoli_NCTC13441_ST131_genome_aln_w_alt_allele_unmapped.filtered_polymorphic_sites.fasta", type = "DNA")
Ecoli_131_var <- read.dna("/Users/tifwan/Desktop/flux_mount_gl/Project_IMPALA/old_Sequence_data/output_files/2024_01_12_IMPALA_variant_calling/Ecoli/EcoliST131/2024_02_29_15_04_03_core_results/gubbins/2024_02_29_15_04_03_Ecoli_NCTC13441_ST131_genome_aln_w_alt_allele_unmapped.filtered_polymorphic_sites.fasta",format = "fasta", as.matrix=TRUE,as.character = TRUE)
Ecoli_131_tree <- read.tree("/Users/tifwan/Desktop/flux_mount_gl/Project_IMPALA/old_Sequence_data/output_files/2024_01_12_IMPALA_variant_calling/Ecoli/EcoliST131/2024_02_29_15_04_03_core_results/gubbins/iqtree_masked_wga/2024_02_29_15_04_03_Ecoli_NCTC13441_ST131_genome_aln_w_alt_allele_unmapped.treefile")
ggtree(Ecoli_131_tree) + geom_tiplab2()
# NZ_UFZF01000006.1
#10
Ecoli_10 <- read.FASTA("/Users/tifwan/Desktop/flux_mount_gl/Project_IMPALA/old_Sequence_data/output_files/2024_03_21_Ecoli_ST_based_variant_calling/ST10/2024_03_26_14_44_48_core_results/gubbins/2024_03_26_14_44_48_ST10_genome_aln_w_alt_allele_unmapped.filtered_polymorphic_sites.fasta", type = "DNA")
Ecoli_10_var <- read.dna("/Users/tifwan/Desktop/flux_mount_gl/Project_IMPALA/old_Sequence_data/output_files/2024_03_21_Ecoli_ST_based_variant_calling/ST10/2024_03_26_14_44_48_core_results/gubbins/2024_03_26_14_44_48_ST10_genome_aln_w_alt_allele_unmapped.filtered_polymorphic_sites.fasta",format = "fasta", as.matrix=TRUE,as.character = TRUE)
Ecoli_10_tree <- read.tree("/Users/tifwan/Desktop/flux_mount_gl/Project_IMPALA/old_Sequence_data/output_files/2024_03_21_Ecoli_ST_based_variant_calling/ST10/2024_03_26_14_44_48_core_results/gubbins/iqtree_masked_wga/2024_03_26_14_44_48_ST10_genome_aln_w_alt_allele_unmapped.treefile")
ggtree(Ecoli_10_tree) + geom_tiplab()
# IMPALA_45

#1598
Ecoli_1598 <- read.FASTA("/Users/tifwan/Desktop/flux_mount_gl/Project_IMPALA/old_Sequence_data/output_files/2024_03_21_Ecoli_ST_based_variant_calling/ST1598/2024_03_26_14_46_44_core_results/gubbins/2024_03_26_14_46_44_ST1598_genome_aln_w_alt_allele_unmapped.filtered_polymorphic_sites.fasta", type = "DNA")
Ecoli_1598_var <- read.dna("/Users/tifwan/Desktop/flux_mount_gl/Project_IMPALA/old_Sequence_data/output_files/2024_03_21_Ecoli_ST_based_variant_calling/ST1598/2024_03_26_14_46_44_core_results/gubbins/2024_03_26_14_46_44_ST1598_genome_aln_w_alt_allele_unmapped.filtered_polymorphic_sites.fasta", format = "fasta", as.matrix=TRUE,as.character = TRUE)
Ecoli_1598_tree <- read.tree("/Users/tifwan/Desktop/flux_mount_gl/Project_IMPALA/old_Sequence_data/output_files/2024_03_21_Ecoli_ST_based_variant_calling/ST1598/2024_03_26_14_46_44_core_results/gubbins/iqtree_masked_wga/2024_03_26_14_46_44_ST1598_genome_aln_w_alt_allele_unmapped.treefile")
ggtree(Ecoli_1598_tree) + geom_tiplab2()
# IMPALA_1201

#1722
Ecoli_1722 <- read.FASTA("/Users/tifwan/Desktop/flux_mount_gl/Project_IMPALA/old_Sequence_data/output_files/2024_03_21_Ecoli_ST_based_variant_calling/ST1722/2024_03_26_14_48_41_core_results/gubbins/2024_03_26_14_48_41_ST1722_genome_aln_w_alt_allele_unmapped.filtered_polymorphic_sites.fasta", type = "DNA")
Ecoli_1722_var <- read.dna("/Users/tifwan/Desktop/flux_mount_gl/Project_IMPALA/old_Sequence_data/output_files/2024_03_21_Ecoli_ST_based_variant_calling/ST1722/2024_03_26_14_48_41_core_results/gubbins/2024_03_26_14_48_41_ST1722_genome_aln_w_alt_allele_unmapped.filtered_polymorphic_sites.fasta", format = "fasta",as.matrix = TRUE, as.character = TRUE)
Ecoli_1722_tree <- read.tree("/Users/tifwan/Desktop/flux_mount_gl/Project_IMPALA/old_Sequence_data/output_files/2024_03_21_Ecoli_ST_based_variant_calling/ST1722/2024_03_26_14_48_41_core_results/gubbins/iqtree_masked_wga/2024_03_26_14_48_41_ST1722_genome_aln_w_alt_allele_unmapped.treefile")
ggtree(Ecoli_1722_tree) + geom_tiplab2()
# IMPALA_816

#38
Ecoli_38 <- read.FASTA("/Users/tifwan/Desktop/flux_mount_gl/Project_IMPALA/old_Sequence_data/output_files/2024_03_21_Ecoli_ST_based_variant_calling/ST38/2024_03_26_14_50_41_core_results/gubbins/2024_03_26_14_50_41_ST38_genome_aln_w_alt_allele_unmapped.filtered_polymorphic_sites.fasta", type = "DNA")
Ecoli_38_var <- read.dna("/Users/tifwan/Desktop/flux_mount_gl/Project_IMPALA/old_Sequence_data/output_files/2024_03_21_Ecoli_ST_based_variant_calling/ST38/2024_03_26_14_50_41_core_results/gubbins/2024_03_26_14_50_41_ST38_genome_aln_w_alt_allele_unmapped.filtered_polymorphic_sites.fasta", format = "fasta",as.matrix = TRUE, as.character = TRUE)
Ecoli_38_tree <- read.tree("/Users/tifwan/Desktop/flux_mount_gl/Project_IMPALA/old_Sequence_data/output_files/2024_03_21_Ecoli_ST_based_variant_calling/ST38/2024_03_26_14_50_41_core_results/gubbins/iqtree_masked_wga/2024_03_26_14_50_41_ST38_genome_aln_w_alt_allele_unmapped.treefile")
ggtree(Ecoli_38_tree) + geom_tiplab2()
# outgroup = IMPALA_1194

#46
Ecoli_46 <- read.FASTA("/Users/tifwan/Desktop/flux_mount_gl/Project_IMPALA/old_Sequence_data/output_files/2024_03_21_Ecoli_ST_based_variant_calling/ST46/2024_03_26_14_52_18_core_results/gubbins/2024_03_26_14_52_18_ST46_genome_aln_w_alt_allele_unmapped.filtered_polymorphic_sites.fasta", type = "DNA")
Ecoli_46_var <- read.dna("/Users/tifwan/Desktop/flux_mount_gl/Project_IMPALA/old_Sequence_data/output_files/2024_03_21_Ecoli_ST_based_variant_calling/ST46/2024_03_26_14_52_18_core_results/gubbins/2024_03_26_14_52_18_ST46_genome_aln_w_alt_allele_unmapped.filtered_polymorphic_sites.fasta", format = "fasta",as.matrix = TRUE, as.character = TRUE)
Ecoli_46_tree <- read.tree("/Users/tifwan/Desktop/flux_mount_gl/Project_IMPALA/old_Sequence_data/output_files/2024_03_21_Ecoli_ST_based_variant_calling/ST46/2024_03_26_14_52_18_core_results/gubbins/iqtree_masked_wga/2024_03_26_14_52_18_ST46_genome_aln_w_alt_allele_unmapped.treefile")
ggtree(Ecoli_46_tree) + geom_tiplab2()
# outgroup = IMPALA_1095

#636
Ecoli_636 <- read.FASTA("/Users/tifwan/Desktop/flux_mount_gl/Project_IMPALA/old_Sequence_data/output_files/2024_03_21_Ecoli_ST_based_variant_calling/ST636/2024_03_26_14_54_51_core_results/gubbins/2024_03_26_14_54_51_ST636_genome_aln_w_alt_allele_unmapped.filtered_polymorphic_sites.fasta", type = "DNA")
Ecoli_636_var <- read.dna("/Users/tifwan/Desktop/flux_mount_gl/Project_IMPALA/old_Sequence_data/output_files/2024_03_21_Ecoli_ST_based_variant_calling/ST636/2024_03_26_14_54_51_core_results/gubbins/2024_03_26_14_54_51_ST636_genome_aln_w_alt_allele_unmapped.filtered_polymorphic_sites.fasta", format = "fasta",as.matrix = TRUE, as.character = TRUE)
Ecoli_636_tree <- read.tree("/Users/tifwan/Desktop/flux_mount_gl/Project_IMPALA/old_Sequence_data/output_files/2024_03_21_Ecoli_ST_based_variant_calling/ST636/2024_03_26_14_54_51_core_results/gubbins/iqtree_masked_wga/2024_03_26_14_54_51_ST636_genome_aln_w_alt_allele_unmapped.treefile")
ggtree(Ecoli_636_tree)+geom_tiplab()
# outgroup = IMPALA_238

#69
Ecoli_69 <- read.FASTA("/Users/tifwan/Desktop/flux_mount_gl/Project_IMPALA/old_Sequence_data/output_files/2024_03_21_Ecoli_ST_based_variant_calling/ST69/2024_03_26_14_59_18_core_results/gubbins/2024_03_26_14_59_18_ST69_genome_aln_w_alt_allele_unmapped.filtered_polymorphic_sites.fasta", type = "DNA")
Ecoli_69_var <- read.dna("/Users/tifwan/Desktop/flux_mount_gl/Project_IMPALA/old_Sequence_data/output_files/2024_03_21_Ecoli_ST_based_variant_calling/ST69/2024_03_26_14_59_18_core_results/gubbins/2024_03_26_14_59_18_ST69_genome_aln_w_alt_allele_unmapped.filtered_polymorphic_sites.fasta", format = "fasta",as.matrix = TRUE, as.character = TRUE)
Ecoli_69_tree <- read.tree("/Users/tifwan/Desktop/flux_mount_gl/Project_IMPALA/old_Sequence_data/output_files/2024_03_21_Ecoli_ST_based_variant_calling/ST69/2024_03_26_14_59_18_core_results/gubbins/iqtree_masked_wga/2024_03_26_14_59_18_ST69_genome_aln_w_alt_allele_unmapped.treefile")
ggtree(Ecoli_69_tree) + geom_tiplab()
# outgroup = IMPALA_677

#757
Ecoli_757 <- read.FASTA("/Users/tifwan/Desktop/flux_mount_gl/Project_IMPALA/old_Sequence_data/output_files/2024_03_21_Ecoli_ST_based_variant_calling/ST757/2024_03_26_15_01_22_core_results/gubbins/2024_03_26_15_01_22_ST757_genome_aln_w_alt_allele_unmapped.filtered_polymorphic_sites.fasta", type = "DNA")
Ecoli_757_var <- read.dna("/Users/tifwan/Desktop/flux_mount_gl/Project_IMPALA/old_Sequence_data/output_files/2024_03_21_Ecoli_ST_based_variant_calling/ST757/2024_03_26_15_01_22_core_results/gubbins/2024_03_26_15_01_22_ST757_genome_aln_w_alt_allele_unmapped.filtered_polymorphic_sites.fasta", format = "fasta",as.matrix = TRUE, as.character = TRUE)
Ecoli_757_tree <- read.tree("/Users/tifwan/Desktop/flux_mount_gl/Project_IMPALA/old_Sequence_data/output_files/2024_03_21_Ecoli_ST_based_variant_calling/ST757/2024_03_26_15_01_22_core_results/gubbins/iqtree_masked_wga/2024_03_26_15_01_22_ST757_genome_aln_w_alt_allele_unmapped.treefile")
# outgroup =IMPALA_43
```

# create distance matrices
```{r echo=FALSE, results = FALSE}
Ecoli_131_list <- get_ST_variant_call(Ecoli_131, "NZ_UFZF01000006.1")
get_ST_variant_call(Ecoli_131)
Ecoli_131_mat_vc <- Ecoli_131_list[[1]]
Ecoli_131_mat_long_vc <- Ecoli_131_list[[2]]

Ecoli_10_list <- get_ST_variant_call(Ecoli_10)
Ecoli_10_mat_vc <- Ecoli_10_list[[1]]
Ecoli_10_mat_long_vc <- Ecoli_10_list[[2]]

Ecoli_1598_list <- get_ST_variant_call(Ecoli_1598)
Ecoli_1598_mat_vc <- Ecoli_1598_list[[1]]
Ecoli_1598_mat_long_vc <- Ecoli_1598_list[[2]]

Ecoli_1722_list <- get_ST_variant_call(Ecoli_1722)
Ecoli_1722_mat_vc <- Ecoli_1722_list[[1]]
Ecoli_1722_mat_long_vc <- Ecoli_1722_list[[2]]

Ecoli_38_list <- get_ST_variant_call(Ecoli_38)
Ecoli_38_mat_vc <- Ecoli_38_list[[1]]
Ecoli_38_mat_long_vc <- Ecoli_38_list[[2]]

Ecoli_46_list <- get_ST_variant_call(Ecoli_46)
Ecoli_46_mat_vc <- Ecoli_46_list[[1]]
Ecoli_46_mat_long_vc <- Ecoli_46_list[[2]]

Ecoli_636_list <- get_ST_variant_call(Ecoli_636)
Ecoli_636_mat_vc <- Ecoli_636_list[[1]]
Ecoli_636_mat_long_vc <- Ecoli_636_list[[2]]

Ecoli_69_list <- get_ST_variant_call(Ecoli_69)
Ecoli_69_mat_vc <- Ecoli_69_list[[1]]
Ecoli_69_mat_long_vc <- Ecoli_69_list[[2]]

Ecoli_757_list <- get_ST_variant_call(Ecoli_757)
Ecoli_757_mat_vc <- Ecoli_757_list[[1]]
Ecoli_757_mat_long_vc <- Ecoli_757_list[[2]]

```
# merge with pairwise SNV distance
```{r}
Ecoli_SNV_long <- rbind(Ecoli_131_mat_long_vc,Ecoli_10_mat_long_vc, Ecoli_1598_mat_long_vc,Ecoli_1722_mat_long_vc,Ecoli_38_mat_long_vc, Ecoli_46_mat_long_vc, Ecoli_636_mat_long_vc, Ecoli_69_mat_long_vc, Ecoli_757_mat_long_vc)

```



# pairwise distance histogram 
```{r}
ST <- c(rep("131",nrow(Ecoli_131_mat_long_vc)),rep("10",nrow(Ecoli_10_mat_long_vc)),rep("1598",nrow(Ecoli_1598_mat_long_vc)),rep("1722",nrow(Ecoli_1722_mat_long_vc)),rep("38",nrow(Ecoli_38_mat_long_vc)),rep("46",nrow(Ecoli_46_mat_long_vc)),rep("636",nrow(Ecoli_636_mat_long_vc)),rep("69",nrow(Ecoli_69_mat_long_vc)),rep("757",nrow(Ecoli_757_mat_long_vc)))

ST_mat_long <- rbind(Ecoli_131_mat_long_vc, Ecoli_10_mat_long_vc, Ecoli_1598_mat_long_vc,Ecoli_1722_mat_long_vc,Ecoli_38_mat_long_vc,Ecoli_46_mat_long_vc, Ecoli_636_mat_long_vc, Ecoli_69_mat_long_vc, Ecoli_757_mat_long_vc)
ST_mat_long$ST <- ST

#Ecoli_SNV_long <- merge(Ecoli_SNV_long,msv_pairs, by = c("X1","X2"), all = TRUE) %>%
 # mutate(msv_pair = ifelse(is.na(msv),"No","Yes"))

Ecoli_SNV_long <- ST_mat_long %>%
  mutate(transmission = ifelse(dist <=5, 1,0))

Ecoli_SNV_long %>%
  ggplot() + geom_histogram(aes(x=dist, fill = as.factor(transmission)), alpha=.6) + geom_vline(xintercept=c(10,15), linetype = "dashed") + scale_x_continuous(limit = c(0,50)) + facet_wrap(.~ST, scales = "free_y") + labs(fill = "SNV<=5")

ggsave("/Users/tifwan/University of Michigan Dropbox/Tiffany Wan/IMPALA-2/Manuscripts/Decriptive genomic epi paper/2025_04_26/supplemental_5.pdf")
Ecoli_SNV_long <- Ecoli_SNV_long %>%
  select(-ST)

```
# 
```{r}
# mat_long <- rbind(Ecoli_131_mat_long, Ecoli_10_mat_long, Ecoli_1598_mat_long,Ecoli_1722_mat_long, Ecoli_38_mat_long, Ecoli_46_mat_long, Ecoli_636_mat_long, Ecoli_69_mat_long, Ecoli_757_mat_long)

ST_demo_sub1 <- ST_demo %>%
  select(-c("site_location","isolate_numbers","external_participant_id","participant","date_of_admission","bed_num","ward","id_referring_subject","use_hhid_referring_subj","ass_new_hhid","rsns_for_hospitilization","othr_cr_area","participant_age","othr_rsn_hos","sele_hos_name","spec_othr_rsn_hos_2","othr_cr_area_2","spec_num_dys","Prep method","whch_cntry_trvalled_4___4","Sequencing plate","pst_thre_mnth_hos_any_rsn","fo_hw_mny_days_hosp","hw_mny_days_fr_disc_enr","spec_num_dys_frm_dis_enrl","wht_cr_area_hospitilized","fo_hw_mny_days_hosp_2","othr_rsns_travelling","hsehld_id","num_cmpl_dt_antibio_enrll","spec_num_dys_anti_tkn","no_cmpltn_dt_anti_knwn","pst_thremnth_antbio_use","num_dys_antibio_tkn","pst6mnth_mbr_vst_cl_mdcl_9","pst6mnth_mbr_vst_hs_nt_mc9","pst6mnth_mbr_vst_cl_nt_mc9")) %>%
  select(-contains(c("whch_cntry_trvalled","hospitalizations","fo_hw_mny_days_hosp","rsns_for_travel","no_dys_thr","rsns_for_hospitilization","wht_cr_area_hospitilized", "hw_mny_days_fr_disc_enr","trvl_episodes","spec_antibio_nam","no_of_antibiotics", "whr_antibio_obtnd","spec_num_dys_fr_dis_enrl","num_cmpl_dt_antibio_enrl", "spec_num_dys_anti_tkn","no_cmpltn_dt_anti_knwn","nam_antibio","num_dys_antibio_tkn","hw_mny_dys_out_ctry","othr_rsn_hos","hw_mny_dys_rtn_enrl","spec_num_dys","pst6mnth_mbr_vst_hs_mdcl","household_members","othr","cntry","num_days","sele_clinic_nam")))

ST_demo_sub1[,c("hw_ftn_visit_hos_med_care","hw_oftn_vst_clinic_med_cr","hw_oftn_vst_clnc_nt_med_cr","hw_oftn_vst_hs_not_medcr","grow_crps","tend_livstck_home","tend_pltry_home","tend_swine_home","tend_livstck_farm","tend_pltry_farm","tend_swine_farm","catch_handle_fish","prep_fish_chickn_consup","handle_trt_disp_waste","livestock_cattle_goats","chicken","turkeys_ducks","horses_donkeys","swine","grw_tend_crps_9","tend_livestock_hme_9","tend_poultry_hme_9","tend_swine_hme_9","tend_livestock_farm_9","tend_poultry_farm_9")]  <- apply(ST_demo_sub1[,c("hw_ftn_visit_hos_med_care","hw_oftn_vst_clinic_med_cr","hw_oftn_vst_clnc_nt_med_cr","hw_oftn_vst_hs_not_medcr","grow_crps","tend_livstck_home","tend_pltry_home","tend_swine_home","tend_livstck_farm","tend_pltry_farm","tend_swine_farm","catch_handle_fish","prep_fish_chickn_consup","handle_trt_disp_waste","livestock_cattle_goats","chicken","turkeys_ducks","horses_donkeys","swine","grw_tend_crps_9","tend_livestock_hme_9","tend_poultry_hme_9","tend_swine_hme_9","tend_livestock_farm_9","tend_poultry_farm_9")], 2, function(x){
  r = ifelse(grepl("Not once",x)|grepl("Unknown",x)|grepl("Less than 1 day a week",x)|grepl("Other",x)|grepl("Never",x),0,1)
  return(r)
})
ST_demo_sub2 <- ST_demo_sub1 %>%
  #select(Sample,ST, age1,hhid,location,participant_type,participant_type1,social,lockdown) %>%
  mutate(region = case_when(location %in% c("Bokaa Clinic","Boseja Clinic","Deborah Retief Memorial Hospital")~"Mochudi",
                            location %in% c("Borakalalo Clinic","Phuthadikobo Clinic","Scottish Livingstone Referral Hospital") ~"Molepolele",
                            location %in% c("Julia Molefhe Clinic","Nkoyaphiri Clinic","Princes Marina Hospital") ~"Gaborone"))
#Ecoli_meta_match <- Ecoli_meta_match[,c("age1","hhid_match","location","participant_type","participant_type1","social_no_hh_match")]
```


```{r}
# merge and matke sure the two data frames are in the same order
Ecoli_meta_long1 <- merge(Ecoli_SNV_long , ST_demo_sub2, by.x ="X1" ,by.y = "Sample", all.x = TRUE) %>%
  arrange(X1, X2)
Ecoli_meta_long2 <- merge(Ecoli_SNV_long , ST_demo_sub2, by.x ="X2" ,by.y = "Sample", all.x = TRUE) %>%
  arrange(X1, X2)



Ecoli_meta_match <- Ecoli_meta_long1

Ecoli_meta_match <- Ecoli_meta_match %>%
  select(-collection_date) %>%
  filter((X1%in%ST_demo_sub2$Sample) & (X2%in%ST_demo_sub2$Sample))
for (row in 1:nrow(Ecoli_meta_match)){
  for (col in 7:ncol(Ecoli_meta_match)){
    var <- colnames(Ecoli_meta_match)[col]
    p1 <- Ecoli_meta_match$X1[row]
    p2 <- Ecoli_meta_match$X2[row]
    if (var%in% c("hw_ftn_visit_hos_med_care","hw_oftn_vst_clinic_med_cr","hw_oftn_vst_clnc_nt_med_cr","hw_oftn_vst_hs_not_medcr","cheese1___2","cheese1___3","cow_milk___1","cow_milk___2","cow_milk___3","cow_milk___5","eggs1___1","eggs1___2","eggs1___3","eggs1___4","eggs1___5","frsh_fruit___1","frsh_fruit___2","frsh_fruit___4","frsh_fruit___5","frsh_veg___1","frsh_veg___2","frsh_veg___3","frsh_veg___4","frsh_veg___5","goat_milk___1","goat_milk___2","goat_milk___3","goat_milk___5","is_par_hos_patient","main_src_drnkn_watr___1","main_src_drnkn_watr___10","main_src_drnkn_watr___11","main_src_drnkn_watr___12","main_src_drnkn_watr___13","main_src_drnkn_watr___14","main_src_drnkn_watr___2","main_src_drnkn_watr___3","main_src_drnkn_watr___4","main_src_drnkn_watr___5","main_src_drnkn_watr___6","main_src_drnkn_watr___9","main_src_wtr_nt_drnkn___1","main_src_wtr_nt_drnkn___10","main_src_wtr_nt_drnkn___11","main_src_wtr_nt_drnkn___12","main_src_wtr_nt_drnkn___13","main_src_wtr_nt_drnkn___14","main_src_wtr_nt_drnkn___2","main_src_wtr_nt_drnkn___3","main_src_wtr_nt_drnkn___4","main_src_wtr_nt_drnkn___5","main_src_wtr_nt_drnkn___6","main_src_wtr_nt_drnkn___9","meat___1","meat___2","meat___3","meat___4","meat___5","poultry___1","poultry___2","poultry___3","poultry___4","poultry___5","seafood___2","seafood___3","seafood___4","seafood___6","yogurt___2","grow_crps","tend_livstck_home","tend_pltry_home","tend_swine_home","tend_livstck_farm","tend_pltry_farm","tend_swine_farm","catch_handle_fish","prep_fish_chickn_consup","handle_trt_disp_waste","livestock_cattle_goats","chicken","turkeys_ducks","horses_donkeys","swine","grw_tend_crps_9","tend_livestock_hme_9","tend_poultry_hme_9","tend_swine_hme_9","tend_livestock_farm_9","tend_poultry_farm_9")){
      if (is.na(ST_demo_sub2[[var]][which(ST_demo_sub2$Sample==p1)])|is.na(ST_demo_sub2[[var]][which(ST_demo_sub2$Sample==p2)])){
        Ecoli_meta_match[row,col] <- NA}
      else if (ST_demo_sub2[[var]][which(ST_demo_sub2$Sample==p1)]== ST_demo_sub2[[var]][which(ST_demo_sub2$Sample==p2)]){
        Ecoli_meta_match[row,col] <- "Yes"
      }
      else {
        Ecoli_meta_match[row,col] <- "No"
      }
      }
    else if(is.na(ST_demo_sub2[[var]][which(ST_demo_sub2$Sample==p1)])|is.na(ST_demo_sub2[[var]][which(ST_demo_sub2$Sample==p2)])){
      Ecoli_meta_match[row,col] <- NA
    }
    
    else if (ST_demo_sub2[[var]][which(ST_demo_sub2$Sample==p1)]== ST_demo_sub2[[var]][which(ST_demo_sub2$Sample==p2)]){
      Ecoli_meta_match[row,col] <- ST_demo_sub2[[var]][which(ST_demo_sub2$Sample==p1)]
    }

    
    else{
       Ecoli_meta_match[row,col] <- "No"
    } 
  }
}

# 
# for (row in 1:nrow(Ecoli_meta_long1)){
#   for (col in 11:ncol(Ecoli_meta_long1)){
#     if ( is.na(Ecoli_meta_long1[row,col])|is.na(Ecoli_meta_long2[row,col])){
#       Ecoli_meta_match[row,col] <- NA
#     }
#     else if ( Ecoli_meta_long1[row,col]==Ecoli_meta_long2[row,col]){
#       Ecoli_meta_match[row,col] <- Ecoli_meta_long1[row,col]
#     }
#     else{
#        Ecoli_meta_match[row,col] <- "No"
#     } 
#   }
# }

# rename some of the variables
Ecoli_meta_match <- Ecoli_meta_match %>%
  mutate(hhid_match = ifelse(hhid!="No","Yes",hhid),
         social_no_hh_match = case_when(social!="No" & is.na(hhid)~"Yes",
                                        social!="No" & hhid=="No"~"Yes",
                                        TRUE ~"No")) %>%
  
  filter(!is.na(lockdown))

Ecoli_meta_match <- Ecoli_meta_match %>%
  filter(!ST %in% c("-","No"))
table(Ecoli_meta_match$ST)

# # filter out pairs from the same participants and with the same ST
# Ecoli_distmat_var_long1 <- Ecoli_distmat_var_long1 %>%
#   filter(!(participant=="Yes" & ST =="Yes"))
```

```{r}
# unorder variables, and set refernece value at 'No'

Ecoli_meta_match[,10:ncol(Ecoli_meta_match)] <- apply(Ecoli_meta_match[,10:ncol(Ecoli_meta_match)], 2, function(x){
  x <- factor(x, ordered = FALSE)
  if ("No" %in% x){
    x <- relevel(x, ref = "No")
  }
  
})

```

# Same household pairwise distance
```{r}
Ecoli_meta_match %>%
  filter(hhid_match=="Yes" & lockdown!="No") %>%
  ggplot() + geom_bar(aes(x = dist, ))+ scale_x_continuous(limits = c(-1,30)) + geom_vline(xintercept =5, color =5) + theme_bw() +
    theme(
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      axis.text.x = element_text(angle = 30, vjust =0),
      legend.position = "bottom")

ggsave("/Users/tifwan/University of Michigan Dropbox/Tiffany Wan/IMPALA-2/Manuscripts/Decriptive genomic epi paper/2025_04_26/figure5a.pdf", width = 7, height = 4)

table(Ecoli_meta_match[Ecoli_meta_match$lockdown!="No",]$hhid_match, Ecoli_meta_match[Ecoli_meta_match$lockdown!="No",]$dist<=5)
# len_na <- apply(Ecoli_meta_match, 1, function(x){
#   sum(!is.na(x))})
#Ecoli_meta_match <- Ecoli_meta_match[,c("age1","hhid_match","location","participant_type","participant_type1","social_no_hh_match")]
hh_pair_df <- Ecoli_meta_match %>%
  filter(hhid_match=="Yes" & lockdown!="No")
```

```{r}
covar <- colnames(Ecoli_meta_match[7:ncol(Ecoli_meta_match)])
#covar <- c("loc_match2", "region_match2", "lockdown_match1", "social_no_hh_match", "hhid_match","loc_type_match2","age_match")

remove_var <- c("hhid","social")
var_ls <- sapply(Ecoli_meta_match, function(x){
  unique(x)
})
for (i in names(var_ls)){
  if (length(var_ls[[i]])==1 | length(setdiff(var_ls[[i]],c("No","0","NA")))<=1){
    remove_var <- c(remove_var, i)
  }
}


remove_var <- remove_var [remove_var!="social_no_hh_match"]
covar <- covar[!covar %in% remove_var]

msv_model_ls <- lapply(covar, function(inde_var){
  msv_model <- glm(transmission ~ get(inde_var), data = Ecoli_meta_match, family= binomial(link = "logit"))
})

# Collapse models into dataframe for easy inspection
msv_models_df <- data.frame()
  for(j in 1:length(covar)){
    msv_models_df <- rbind(msv_models_df,
                       data.frame(var = covar[j], # the covariates
                                as.data.frame(coef(summary(msv_model_ls[[j]])))%>%
                                              rownames_to_column(var="covariates"))) # the model fit
  }

msv_models_df <- msv_models_df %>%
  mutate(threshold_c = case_when(Pr...z..<0.05 ~ "P - value < 0.05",
                                 Pr...z..> 0.05 & Pr...z..<0.2 ~ "0.05 < P-value < 0.2",
                                 Pr...z..>0.2 ~ "P-value > 0.2"),
         odds = exp(Estimate)) # whether p-value is below below 0.05, or 0.2 


```

# test if pre-post lockdown have difference in transmission
```{r}

lockdown_pair_df <- Ecoli_meta_match [Ecoli_meta_match $lockdown!="No",]
table(lockdown_pair_df$ST)
table(lockdown_pair_df$lockdown)
table(lockdown_pair_df$transmission, lockdown_pair_df$lockdown)
chisq.test(as.factor(lockdown_pair_df$transmission), lockdown_pair_df$lockdown, correct = T)

#length(unique(lockdown_pair_df$hhid[lockdown_pair_df$participant_live_same_hh=="Yes"]))
```





```{r}
Ecoli_meta_match$transmission <- as.factor(Ecoli_meta_match$transmission)
Ecoli_meta_match %>%
    filter(!is.na(lockdown)) %>%
   filter(hhid=="No") %>%
    ggplot(aes(x = dist)) +
    geom_density(aes(fill = lockdown),alpha = 0.6, color = NA, adjust=0.3) +
    scale_fill_discrete(labels = c(
      "pre" = paste("Pre-lockdown ( N=", length(which(Ecoli_meta_match$lockdown=="pre")),")"),
      "post" = paste("Post-lockdown  ( N=", length(which(Ecoli_meta_match$lockdown=="post")),")"))) +
    labs(x = "SNP", y = "%") +
    scale_x_continuous(limits = c(0, 50), breaks = seq(0, 50, 10)) 
#ggsave("~/Dropbox (University of Michigan)/IMPALA-2/Manuscripts/Decriptive genomic epi paper/Figures_with_3rd_batch/same_ST_lockdown_density.pdf", height = 5, width = 7)
```




```{r}


Ecoli_meta_match <- Ecoli_meta_match %>%
  filter(!is.na(lockdown))
for (i in c("131","10","69","1722","757","1598")){
  for( var in c("region","participant_type1","location")){
    g <- create_graph(Ecoli_meta_match,c(i))
  lay <- create_layout(g,layout = 'fr')

  print(
    ggraph(lay)  + geom_edge_link(aes(color = lockdown),alpha=.5) + geom_node_point(aes(color = get(var))) +scale_edge_color_manual(values= c("pre" = "#80CDC1","post"="#D01C8B","No"="#FDE725FF")) +theme_void() + labs(title = paste("ST:",i)) +scale_color_viridis_d()
  )
  ggsave(paste0("~/Dropbox (University of Michigan)/IMPALA-2/Manuscripts/Decriptive genomic epi paper/Figures_with_3rd_batch/network/network_",i,"_",var,".pdf"))  

  }
}

# all ST
Ecoli_meta_match <- Ecoli_meta_match %>%
  filter(!is.na(lockdown))


g <- create_graph(Ecoli_meta_match,c("131","10","69","1722","757","1598"))
lay <- create_layout(g,layout = 'nicely')

ggraph(lay)  + geom_edge_link(aes(color = lockdown),alpha=.5) + geom_node_point(aes(color = ST),alpha=.8) +scale_edge_color_manual(values= c("pre" = "#80CDC1","post"="#D01C8B","No"="#FDE725FF")) +theme_void() +scale_color_viridis_d() + facet_wrap(lockdown~.)

#ggsave("~/Dropbox (University of Michigan)/IMPALA-2/Manuscripts/Decriptive genomic epi paper/Figures_with_3rd_batch/pre_post_ST_network.pdf", height = 5, width = 10)  




```

```{r}
create_graph <- function(dist_mat_long, condition){
  #test data
   # dist_mat_long = Ecoli_meta_match1
   # hclust_df = hclust_Ecoli
   # condition = c("131")
   # n = 10
   # height =10
  
  trans_edge <- dist_mat_long %>%
  #filter(ST_match=="Yes") %>%
  #mutate(dist = as.numeric(dist)) %>%
  filter(transmission==1) %>% 
  filter(ST %in% condition) %>%
  select(X1,X2,lockdown) #%>%
  #filter(lockdown!="No")

  pt_node <- data.frame(nodes = unique(c(trans_edge$X1,trans_edge$X2))) %>%
    merge(ST_demo[,c("Sample","ST","lockdown","participant_type","participant_type1","location","participant_live_same_hh","referred_by_another")], by.x = "nodes", by.y = "Sample")

  # create graph from data frames
  g = tbl_graph(pt_node,trans_edge,  directed = FALSE) %>% activate(nodes)%>%
    
    mutate(group = group_fast_greedy(),
           region = case_when(location %in% c("Bokaa Clinic","Boseja Clinic","Deborah Retief Memorial Hospital")~"Mochudi",
                            location %in% c("Borakalalo Clinic","Phuthadikobo Clinic","Scottish Livingstone Referral Hospital") ~"Molepolele",
                            location %in% c("Julia Molefhe Clinic","Nkoyaphiri Clinic","Princes Marina Hospital") ~"Gaborone"))
  
  return(g)
}

for (i in c("131","10","69","1722","757","1598")){
  for( var in c("region","location","participant_type1","participant_type","participant_live_same_hh")){
    g <- create_graph(Ecoli_meta_match,c(i))
  lay <- create_layout(g,layout = 'fr')

  print(
    ggraph(lay)  + geom_edge_link(aes(color = lockdown),alpha=.5) + geom_node_point(aes(color = get(var)),alpha=.8) +scale_edge_color_manual(values= c("pre" = "#2A788EFF","post"="#440154FF","No"="#FDE725FF")) +theme_void() + labs(title = paste("ST:",i)) #+scale_color_manual(values= c("Hospital patient" = "red","Community/Clinic"="yellow"))
  )
  
  }
}

```

# pre-post lockdown model

# subsetting pre and post models
```{r}
# excluding 1722
# subsetting data w/o 1722
pre_Ecoli_meta <- Ecoli_meta_match %>%
  filter(lockdown=="pre") %>%
  filter(ST!="1722")

post_Ecoli_meta <- Ecoli_meta_match %>%
  filter(lockdown=="post") %>%
  filter(ST!="1722")

table(Ecoli_meta_match$lockdown)

pre_Ecoli_meta_all <- Ecoli_meta_match %>%
  filter(lockdown=="pre")

table(pre_Ecoli_meta_all$location)

post_Ecoli_meta_all <- Ecoli_meta_match %>%
  filter(lockdown=="post")
```
```{r}
for(lockdown_v in c("pre","post")){
  for (var in  c("hhid_match","location","participant_type", "region","social_no_hh_match")){
  if(var =="hhid_match"){
    # df <- Ecoli_meta_match %>%
    #   filter(lockdown =="pre") %>%
    #   filter(!is.na(hhid_match)) %>%
    #   group_by(ST, transmission) %>%
    #   mutate(ST_count = n()) %>%
    #   group_by(ST, get(var), transmission) %>%
    #   summarise(var_prop = n()/ST_count) %>%
    #   distinct()
    
      g <- Ecoli_meta_match %>%
        filter(lockdown ==lockdown_v) %>%
      filter(!is.na(hhid_match)) %>%
       filter(get(var)!="No") %>%
         ggplot() + geom_bar(aes(x =get(var), fill=transmission)) + facet_wrap(.~ST, scales = "free_y")  + labs(title = var) + theme(axis.text.x = element_text(angle = 45,hjust =1))
      print(g)
  #    g <- df %>%
  # filter(!is.na(hhid_match)) %>%
  # ggplot() + geom_jitter(aes(y = transmission, x = var_prop, color = ST))+ facet_wrap(.~ST, ncol = 5) + labs(title = var)
  # print(g) 
  }
  else{
     g <- Ecoli_meta_match %>%
        filter(lockdown ==lockdown_v) %>%
      filter(hhid_match!="Yes"| is.na(hhid_match)) %>%
       filter(get(var)!="No") %>%
         ggplot() + geom_bar(aes(x =get(var), fill=transmission)) + facet_wrap(.~ST, scales = "free_y")  + labs(title = paste0(var," ",lockdown_v,"-lockdown")) + theme(axis.text.x = element_text(angle = 45,hjust =1))
      print(g)
    
  }
 
}
}

post_match_1722 <- post_Ecoli_meta_all %>%
  filter(ST=="1722" & participant_type=="Hospital patient") %>%
  select(X1, X2, transmission, participant_type,location)
```

```{r}
### SNV distance sensitivity analysis
# create covar
pre_covar <- colnames(Ecoli_meta_match[7:ncol(Ecoli_meta_match)])
pre_covar <- pre_covar[pre_covar!="lockdown"]
remove_var <- c("hhid","social")
pre_covar <- pre_covar[!pre_covar %in% remove_var]

# create covar
post_covar <- colnames(Ecoli_meta_match[7:ncol(Ecoli_meta_match)])
post_covar <- post_covar[post_covar!="lockdown"]
remove_var <- c("hhid","social")
post_covar <- post_covar[!post_covar %in% remove_var]
```

```{r}
## might need to try brms - bayesian stuff because they are just not converging

lockdown_var <- c("hhid_match","location","participant_type", "region","social_no_hh_match")
# # for all the pre-covid variables
# pre_model_df <- data.frame()
# for (var in lockdown_var){
#   new_df <- fit_mixed_log_model(df = pre_Ecoli_meta, dep_var = "transmission", indep_var = var)
#   pre_model_df <- rbind(pre_model_df, new_df)
# }
# 
# post_model_df <- data.frame()
# for (var in lockdown_var){
#   new_df <- fit_mixed_log_model(df = post_Ecoli_meta, dep_var = "transmission", indep_var = var)
#   post_model_df <- rbind(post_model_df, new_df)
# }

# for all the pre-covid variables
pre_model_df_all <- data.frame()
for (var in lockdown_var){
  if(var=="hhid_match"){
    new_df <- fit_mixed_log_model(df = pre_Ecoli_meta_all, dep_var = "transmission", indep_var = var)
  }
 else{
   subdf <- pre_Ecoli_meta_all %>%
     filter(hhid_match!="Yes"| is.na(hhid_match))
   new_df <- fit_mixed_log_model(df = subdf, dep_var = "transmission", indep_var = var)
 }
  pre_model_df_all <- rbind(pre_model_df_all, new_df)
}

post_model_df_all <- data.frame()
for (var in lockdown_var){
  if(var=="hhid_match"){
    new_df <- fit_mixed_log_model(df = post_Ecoli_meta_all, dep_var = "transmission", indep_var = var)
  }
  else{
    subdf <- post_Ecoli_meta_all %>%
      filter(hhid_match!="Yes"| is.na(hhid_match))
    new_df <- fit_mixed_log_model(df = subdf, dep_var = "transmission", indep_var = var)
  }
  
  post_model_df_all <- rbind(post_model_df_all, new_df)
}

# location is unable to converge so use bayesian method instead to run model
```
# Bayesian fixed effect transmission risk for location

# this is just a temporary fix, gotta do the trycatch thing
```{r}
# run the mixed effect model first (becasue var_cat in the bayesian model fit depends on its output)
# location
# # pre-lockdown
# bayes_model_fit <- function(df, dep_var,indep_var, var_cat){
# #   dep_var = "transmission"
# # indep_var = "location"
# # var_cat <- pre_model_df$var_cat[pre_model_df$var==indep_var]
# # df = pre_Ecoli_meta_all
# sub_df <- df %>%
#   filter(!is.na(get(indep_var)))
# ### Bayesian model ###
# model <- stan_glmer(get(dep_var) ~ get(indep_var) + (1 | X1) + (1 | X2) , 
#                data = sub_df,
#                family = binomial(link = "logit"),
#                iter = 2000,
#                warmup = 1000,
#                chains =4,
#                control = list(adapt_delta = 0.95))
# # estimates
# estimates <- fixef(model)[-1] # remove intercept term
# # p-value
# posterior <- as.matrix(model)
# p_val <- c()
# for (loc in var_cat){
#   p_val = c(p_val,setNames(mean(posterior[,loc]>0),loc))
#   # print(p_val)
# }
# # odds
# odds <- exp(estimates) 
# # confidence interval
# lowerCI = c()
# upperCI = c()
# for (loc in var_cat){
#    lowerCI = c(lowerCI, setNames(posterior_interval(model,prob = 0.95)[loc,1],loc))
#    upperCI = c(upperCI, setNames(posterior_interval(model,prob = 0.95)[loc,2],loc))
# }
# 
# bayes_model_df <- data.frame(
#   var = "location",
#   var_cat = var_cat,
#   est = estimates,
#   p.value = p_val,
#   odds = odds,
#   lowerCI = lowerCI,
#   upperCI = upperCI
# )
# return(bayes_model_df)
# }
# 
# #   dep_var = "transmission"
# # indep_var = "location"
# # var_cat <- pre_model_df$var_cat[pre_model_df$var==indep_var]
# # df = pre_Ecoli_meta_all
# var_cat <- pre_model_df_all$var_cat[pre_model_df_all$var=="location"]
# pre_baye_model_df <- bayes_model_fit(df = pre_Ecoli_meta_all, dep_var = "transmission",indep_var = "location", var_cat = var_cat)
# var_cat <- post_model_df_all$var_cat[post_model_df_all$var=="location"]
# post_baye_model_df <- bayes_model_fit(df = post_Ecoli_meta_all, dep_var = "transmission",indep_var = "location", var_cat = var_cat)
# 
# # add the bayesian model outputs
# pre_model_df_all <- pre_model_df_all %>%
#   filter(var!="location") %>%
#   select(-fit) %>%
#   rbind(pre_baye_model_df)
# 
# post_model_df_all <- post_model_df_all %>%
#   filter(var!="location") %>%
#   select(-fit) %>%
#   rbind(post_baye_model_df)
# 
# # 
# # # post-lockdown
# # post_sub_df <- post_Ecoli_meta_all %>%
# #   filter(!is.na(get(indep_var)))
# # ### Bayesian model ###
# # post_stan_model <- stan_glmer(get(dep_var) ~ get(indep_var) + (1 | X1) + (1 | X2) , 
# #                data = sub_df,
# #                family = binomial(link = "logit"),
# #                iter = 2000,
# #                warmup = 1000,
# #                chains =4,
# #                control = list(adapt_delta = 0.95))
# # 
# # fixef(post_stan_model)

```


## pre-post lockdown proportions
```{r}
library(cowplot)
# lockdown_var <- intersect(pre_covar,post_covar)
# lockdown_var <- lockdown_var[!lockdown_var%in%c("meat___6","poultry___6","eggs1___6","cow_milk___6","goat_milk___6","yogurt___6","cheese1___6","frsh_fruit___6","frsh_veg___6","referred_by_another","participant_live_same_hh")]
Ecoli_pre_df <- Ecoli_meta_match %>%
  filter(lockdown=="pre")


# plot stats without 1722
pre_summary <- create_pair_summary(stats_df = pre_model_df_all, pairiwise_var_df = pre_Ecoli_meta_all, var_v = lockdown_var)
post_summary <- create_pair_summary(stats_df = post_model_df_all, pairiwise_var_df = post_Ecoli_meta_all, var_v = lockdown_var)

lockdown_df <- pre_summary[[2]] %>%
  mutate(Lockdown = "Pre-Lockdown") %>%
  merge(pre_summary[[1]][,c("group","variables","p.value","odds", "upperCI","lowerCI")],by = c("group","variables")) %>%
  rbind(post_summary[[2]] %>% mutate(Lockdown = "Post-Lockdown") %>% merge(post_summary[[1]][,c("group","variables","p.value","odds", "upperCI","lowerCI")],by = c("group","variables")))

lockdown_df <- lockdown_df%>% distinct() %>%
  filter(!is.na(p.value)|variables=="No")

lockdown_df <- lockdown_df %>%
  group_by(group, Lockdown,Pair_SNV_bool) %>%
  mutate(uniq_var_n = length(unique(variables))) %>%
  filter(uniq_var_n>1 )%>%
  #group_by(group,  Lockdown) %>%
  mutate(p_adjust = p.adjust(p.value, method = "bonferroni"),
         FDR = p.adjust(p.value, method ="BH"))
#lockdown_df$variables <- factor(lockdown_df$variables, levels=c("No","Yes","Bokaa Clinic","Borakalalo Clinic","Boseja Clinic","Julia Molefhe Clinic","Nkoyaphiri Clinic","Phuthadikobo Clinic","Princes Marina Hospital","Scottish Livingstone Referral Hospital","Deborah Retief Memorial Hospital","Gaborone","Clinic patient","Mochudi","Molepolele","Community adult","Community child","Hospital patient","Community/Clinic","Hospital","age <= 5","age 6 to 18","age > 18"))
# relevel the variable for plotting
lockdown_df$Lockdown <- factor(lockdown_df$Lockdown, levels=c("Pre-Lockdown","Post-Lockdown"))

g <- lockdown_df %>%
  group_by(group, variables,Lockdown) %>%
  mutate(prop = Freq/sum(Freq)) %>%
  filter(Pair_SNV_bool==1)  %>%
  mutate(`p<0.05` = ifelse(p.value<=0.05 & odds >=1,"*",""),
         `FDR<0.05` = ifelse(FDR<0.05, "*","")) %>%
  mutate(`Shared risk` = ifelse(variables=="No","No","Yes")) %>%
  filter(group!="participant_type1") %>%
  ggplot() + geom_col(aes(x = variables, y= prop*100, fill = `Shared risk`)) +geom_text(aes(x = variables,label = `p<0.05`),y = 1, size = 7)  + labs( y = "% Transmission pairs with shared demographics",x = "Demographics")+ scale_fill_viridis_d(name = "Shared\ndemographics", label =c("0"="No","1"="Yes")) +theme_void()+ theme_bw() + theme(axis.text.x = element_text(angle = 40,hjust = 1), panel.grid.major.x = element_blank(), panel.grid.minor = element_blank()) + facet_wrap(Lockdown~group, scales = "free",ncol=8) # + ggh4x::facet_grid2(Lockdown~group,scales = "free_x", independent = "x")
g
#ggsave("/Users/tifwan/Dropbox (University of Michigan)/IMPALA-2/Manuscripts/Decriptive genomic epi paper/Figures/variant_lockdown/bar_wo_1722.pdf",  height = 30,width = 15)

#lockdown_df$variables <- factor(lockdown_df$variables, levels=c("No","Princes Marina Hospital","Nkoyaphiri Clinic","Julia Molefhe Clinic","Deborah Retief Memorial Hospital","Boseja Clinic","Bokaa Clinic","Scottish Livingstone Referral Hospital","Phuthadikobo Clinic","Borakalalo Clinic","Gaborone","Mochudi","Molepolele","Yes","Clinic","Hospital","age <= 5","age 6 to 18","age > 18"))

lockdown_df %>%
  group_by(group, variables,Lockdown) %>%
  mutate(prop = Freq/sum(Freq)) %>%
  filter(Pair_SNV_bool==1) %>%
  #filter(Lockdown=="Post-Lockdown") %>%
  filter(!is.na(variables) & variables!="No") %>%
  #filter(group != "location") %>%
  ggplot(aes(x = odds, y = variables)) + 
 geom_vline(aes(xintercept = 1), size = .25, linetype = "dashed", color = "red") +
geom_errorbarh(aes(xmax = upperCI, xmin = lowerCI), size = .5, height = 
.2, color = "gray50") +
   geom_point(x = 0, y = 1,alpha=0) + 
 geom_point(size = 3.5, color = "orange") +
 theme_bw() + ggh4x::facet_grid2(group~Lockdown, scales = "free", independent = "x") 
#ggsave("/Users/tifwan/Dropbox (University of Michigan)/IMPALA-2/Manuscripts/Decriptive genomic epi paper/Figures/variant_lockdown/CI_error_wo_1722.pdf", width = 10,height = 9)
# 
# ## plots including 1722
# pre_summary <- create_pair_summary(stats_df = pre_test_df_all, pairiwise_var_df = pre_Ecoli_meta_all, var_v = lockdown_var)
# post_summary <- create_pair_summary(stats_df = post_test_df_all, pairiwise_var_df = post_Ecoli_meta_all , var_v = lockdown_var)
# 
# lockdown_df <- pre_summary[[2]] %>%
#   mutate(Lockdown = "Pre-Lockdown") %>%
#   merge(pre_summary[[1]][,c("group","variables","p.value","odds", "upperCI","lowerCI")],by = c("group","variables")) %>%
#   rbind(post_summary[[2]] %>% mutate(Lockdown = "Post-Lockdown") %>% merge(post_summary[[1]][,c("group","variables","p.value","odds", "upperCI","lowerCI")],by = c("group","variables")))
# 
# lockdown_df <- lockdown_df %>%
#   group_by(group, Lockdown,Pair_SNV_bool) %>%
#   mutate(uniq_var_n = length(unique(variables))) %>%
#   filter(uniq_var_n>1 )
# 
# #lockdown_df$variables <- factor(lockdown_df$variables, levels=c("No","Yes","Bokaa Clinic","Borakalalo Clinic","Boseja Clinic","Julia Molefhe Clinic","Nkoyaphiri Clinic","Phuthadikobo Clinic","Princes Marina Hospital","Scottish Livingstone Referral Hospital","Deborah Retief Memorial Hospital","Gaborone","Clinic patient","Mochudi","Molepolele","Community adult","Community child","Hospital patient","Community/Clinic","Hospital","age <= 5","age 6 to 18","age > 18"))
# # relevel the variable for plotting
# lockdown_df$Lockdown <- factor(lockdown_df$Lockdown, levels=c("Pre-Lockdown","Post-Lockdown"))
# g <- lockdown_df %>%
#   group_by(group, variables,Lockdown) %>%
#   mutate(prop = Freq/sum(Freq)) %>%
#   filter(Pair_SNV_bool==1) %>%
#   mutate(`p<0.05` = ifelse(p.value<=0.05 & odds >=1,"*","")) %>%
#   mutate(`Shared risk` = ifelse(variables=="No","No","Yes")) %>%
#   filter(group!="participant_type1") %>%
#   ggplot() + geom_col(aes(x = variables, y= prop*100, fill = `Shared risk`)) +geom_text(aes(x = variables,label = `p<0.05`),y = 1, size = 7) + labs( y = "% Transmission pairs with shared demographics",x = "Demographics")+ scale_fill_viridis_d(name = "Shared\ndemographics", label =c("0"="No","1"="Yes")) +theme_void()+ theme_bw() + theme(axis.text.x = element_text(angle = 40,hjust = 1), panel.grid.major.x = element_blank(), panel.grid.minor = element_blank()) + facet_wrap(Lockdown~group, scales = "free",ncol=8) #+ ggh4x::facet_grid2(Lockdown~group,scales = "free_x", independent = "x") 
# g
# #ggsave("/Users/tifwan/Dropbox (University of Michigan)/IMPALA-2/Manuscripts/Decriptive genomic epi paper/Figures/variant_lockdown/bar.pdf",  height = 30,width = 15)
# 
# lockdown_df$variables <- factor(lockdown_df$variables, levels=c("No","Princes Marina Hospital","Nkoyaphiri Clinic","Julia Molefhe Clinic","Deborah Retief Memorial Hospital","Boseja Clinic","Bokaa Clinic","Scottish Livingstone Referral Hospital","Phuthadikobo Clinic","Borakalalo Clinic","Gaborone","Mochudi","Molepolele","Yes","Clinic","Hospital","age <= 5","age 6 to 18","age > 18"))
# 
# lockdown_df %>%
#   group_by(group, variables,Lockdown) %>%
#   mutate(prop = Freq/sum(Freq)) %>%
#   filter(Pair_SNV_bool==1) %>%
#   #filter(Lockdown=="Post-Lockdown") %>%
#   filter(!is.na(variables) & variables!="No") %>%
#   #filter(group != "location") %>%
#   ggplot(aes(x = odds, y = variables)) + 
#  geom_vline(aes(xintercept = 1), size = .25, linetype = "dashed", color = "red") +
# geom_errorbarh(aes(xmax = upperCI, xmin = lowerCI), size = .5, height = 
# .2, color = "gray50") +
#    geom_point(x = 0, y = 1,alpha=0) + 
#  geom_point(size = 2, color = "orange") +
#  theme_bw() + ggh4x::facet_grid2(group~Lockdown, scales = "free", independent = "x") 
# #ggsave("/Users/tifwan/Dropbox (University of Michigan)/IMPALA-2/Manuscripts/Decriptive genomic epi paper/Figures/variant_lockdown/CI_error.pdf",  height = 30,width = 15)
# #ggsave("/Users/tifwan/Dropbox (University of Michigan)/IMPALA-2/Manuscripts/Decriptive genomic epi paper/Figures/variant_lockdown/post_CI_error.pdf", width =10,height = 10)

```

```{r}
lockdown_var <- c("age1", "hhid_match","location","participant_type", "region","social_no_hh_match")
lockdown_df$variables<- factor(lockdown_df$variables, levels=c("No","Yes","Bokaa Clinic","Borakalalo Clinic","Boseja Clinic","Julia Molefhe Clinic","Nkoyaphiri Clinic","Phuthadikobo Clinic","Princes Marina Hospital","Scottish Livingstone Referral Hospital","Deborah Retief Memorial Hospital","Gaborone","Clinic patient","Mochudi","Molepolele","Community adult","Community child","Hospital patient","Community/Clinic","Hospital","age <= 5","age 6 to 18","age > 18"))
g <- lockdown_df %>%
  group_by(group, variables,Lockdown) %>%
  mutate(prop = Freq/sum(Freq)) %>%
  filter(Pair_SNV_bool==1) %>%
  mutate(group1 = group,
         group1 = case_when(group=="age1"~"age",
                           group=="hhid_match"~"household",
                           group =="participant_type"~"participant type",
                           group =="social_no_hh_match"~"social network",
                           TRUE~group),
         facility = case_when(variables == "Bokaa Clinic"~ "Clinic A",
                              variables == "Borakalalo Clinic"~"Clinic B",
                              variables == "Boseja Clinic" ~"Clinic C",
                              variables == "Deborah Retief Memorial Hospital"~ "Hospital A",
                              variables == "Julia Molefhe Clinic" ~ "Clinic D",
                              variables == "Nkoyaphiri Clinic" ~ "Clinic E",
                              variables == "Phuthadikobo Clinic" ~ "Clinic F",
                              variables == "Princes Marina Hospital" ~ "Hospital B",
                              variables == "Scottish Livingstone Referral Hospital" ~ "Hospital C",
                              TRUE ~variables)) %>% 
  mutate(`p<0.05` = ifelse(FDR<0.05 & odds >=1,"*","")) %>%
  mutate(`Shared risk` = ifelse(variables=="No","No","Yes")) %>%
  filter(group!="participant_type1") 
g$facility <- factor(g$facility, levels = c("No","Yes","Clinic A","Clinic B","Clinic C","Clinic D","Clinic E","Clinic F","Hospital A","Hospital B","Hospital C","Gaborone","Clinic patient","Mochudi","Molepolele","Community adult","Community child","Hospital patient","Community/Clinic","Hospital","age <= 5","age 6 to 18","age > 18"))
g %>%
  ggplot() + geom_col(aes(x = variables ,y= prop*100, fill = `Shared risk`)) +geom_text(aes(x = variables,label = `p<0.05`),y = 1, size = 7)  + labs( y = "% Transmission pairs with shared demographics",x = "Demographics")+ 
scale_fill_viridis_d(name = "Shared\ndemographics", label =c("0"="No","1"="Yes")) +theme_void()+ theme_bw() + theme(axis.text.x = element_text(angle = 40,hjust = 1), panel.grid.major.x = element_blank(), panel.grid.minor = element_blank()) + facet_grid(Lockdown~group1, scales = "free") # + ggh4x::facet_grid2(Lockdown~group,scales = "free_x", independent = "x")
#write_xlsx(lockdown_df,"~/Dropbox (University of Michigan)/IMPALA-2/Meta-data/mixed_effect_model_output.xlsx")
#ggsave("/Users/tifwan/Dropbox (University of Michigan)/IMPALA-2/Manuscripts/Decriptive genomic epi paper/Figures/variant_lockdown/2024_12_20_lockdown_risk.pdf", width =10)
#ggsave("/Users/tifwan/University of Michigan Dropbox/Tiffany Wan/IMPALA-2/Manuscripts/Decriptive genomic epi paper/2025_04_26/figure_5a.pdf", width = 12)
```
```{r}
library(rlang)
conditions <- c(
  "participant_type == 'Community child'",
  "hhid_match == 'Yes'"
)
dfs <- c("pre_Ecoli_meta_all","post_Ecoli_meta_all")
trans_df <- data.frame()
for(df in dfs){
  for (cond in conditions) {
  expr <- parse_expr(cond)  # Turn string into an R expression
  subset_df <- get(df) %>%
    filter(!!expr)  # Evaluate it in the context of the data
  new_df <- table(subset_df$hhid,subset_df$transmission) %>%
    as.data.frame() %>%
    mutate(condition = cond,
           lockdown = df)
  trans_df <- rbind(trans_df,new_df)
  }
}

trans_df <- trans_df %>%
  filter(Freq!=0)

trans_df %>%
  filter(Var2==1) %>%
  ggplot(aes(x = Var1, y = Freq, fill = condition), position = "identity") + geom_col()+ facet_wrap(.~lockdown,scales = "free",ncol = 1) 
## table for the different demographics
transmission_demo <- table(pre_Ecoli_meta_all$hhid[pre_Ecoli_meta_all$participant_type=="Community child"],pre_Ecoli_meta_all$transmission[pre_Ecoli_meta_all$participant_type=="Community child"]) %>% as.data.frame() %>%
  mutate(demo = "Community child")

table(post_Ecoli_meta_all$hhid[post_Ecoli_meta_all$participant_type=="Community child"], post_Ecoli_meta_all$transmission[post_Ecoli_meta_all$participant_type=="Community child"]) 

table(pre_Ecoli_meta_all$hhid[pre_Ecoli_meta_all$hhid_match=="Yes"],pre_Ecoli_meta_all$transmission[pre_Ecoli_meta_all$hhid_match=="Yes"])

table(post_Ecoli_meta_all$hhid[post_Ecoli_meta_all$hhid_match=="Yes"],post_Ecoli_meta_all$transmission[post_Ecoli_meta_all$hhid_match=="Yes"])
length(unique(pre_Ecoli_meta_all$hhid))
length(unique(post_Ecoli_meta_all$hhid))
```
```{r}

```

# statistical tests
```{r}
ST_chisq <- ST_demo_sub %>%
  mutate(ST_test = ifelse(MLST=="Other","minor","major"))
chisq.test(table(ST_chisq$ST_test,ST_chisq$lockdown))

df_weekly <- ST_demo  %>%
  mutate(week = floor_date(as.Date(collection_date), unit = "week")) %>%
  group_by(week, location) %>%
  summarise(count = n())

ggplot(df_weekly, aes(x = week, y = count, fill = location)) +
  geom_col() +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %d") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
ST_demo %>%
  mutate(collection_date=as.Date(collection_date)) %>%
  select(collection_date,location) %>%
  ggplot(aes(x = collection_date,fill = location)) + geom_bar()  + scale_x_date(date_breaks = "1 month", date_labels = "%b %d") 

```
# percentage of genetically linked children
```{r}
pre_child <- pre_Ecoli_meta_all %>%
  filter(transmission==1 & participant_type=="Community child") %>%
  select(participant_type, ST, region, hhid_match,lockdown) 

post_child <- post_Ecoli_meta_all %>%
  filter(transmission==1 & participant_type=="Community child")  %>%
  select(participant_type, ST, region, hhid_match,lockdown) 
all_child_prop <- rbind(pre_child, post_child)
         
pre_hos <- pre_Ecoli_meta_all %>%
  filter(transmission==1 & participant_type=="Hospital patient") %>%
  select(participant_type, ST, region, location,lockdown) 

post_hos <- post_Ecoli_meta_all %>%
  filter(transmission==1 & participant_type=="Hospital patient") %>%
  select(participant_type, ST, region, location,lockdown) 

all_hos_prop <- rbind(pre_hos, post_hos)
table(all_hos_prop$region)
nrow(all_hos_prop)
```


